Preprocessing and exploration of the narwhal data for the analysis with smoothSDE

### 2025-01-07 R/preprocessing/narwhals/data_preprocessing.Rmd
R code updates

First preprocessing steps include : 
\begin{itemize}
\item keeping only 2018 data and adding a time since tagging column 
\item converting distance to ship and distance to shore to km and adding an ExpShip column defined as the inverse of the \texttt{Dist\_Ship} column when the narwhals are in line of sight from the boat.\texttt{ExpShip} is a proxy for sound exposure level (SEL).
\item Renaming the "Ind" column to "ID" for clarity and converting it to factors.
\item Converting column "Type" in factor with levels trial and intertrial, and adding two columns "ExpShipT" and "ExpShipI" for exposure to shio during trial and intertrial separately
\item Converting column "Posixct" to POSIXct format for date and time
\end{itemize}

@@

```{=tex}
Anormal values of the speed may indicate outliers to remove from the data.  Here we clearly see some outliers in the histogram.
We thus decide to filter by setting a threshold of $20$ km/h for horizontal speed.
```

@@

ggsave(path=plots_folder,filename="all_narwhals_tracks.png", 
       plot = map,width=10,height=5)

@@

@@

```{r plot coastline and land}
plot_land=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey",col="brown")+
  coord_sf(datum=st_crs(32626))
```

@@

Some points that appear on land on the map don't have distance to shore 0. This may be due to the map inaccuracy.
If we compute the distance to shore based on the map, as done in res_land, we get more points on the land.

@@

cropped_data=dataprep[dataprep$time>12,]
pExpShip=ggplot(cropped_data,aes(time,ExpShip,col=ID))+geom_line(color="black")+
  facet_grid(ID~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to ship")

ggsave(path=plots_folder,"realExpShip_through_time.png", plot=pExpShip)
```

>>>>

First preprocessing steps include : 
\begin{itemize}
\item keeping only 2018 data and adding a time since tagging column 
\item converting distance to ship and distance to shore to km and adding an ExpShip column defined as the inverse of the \texttt{Dist\_Ship} column when the narwhals are in line of sight from the boat.\texttt{ExpShip} is a proxy for sound exposure level (SEL).
\item Renaming the "Ind" column to "ID" for clarity and converting it to factors.
\item Converting column "Type" in factor with levels trial and intertrial, and adding two columns "ExpShipT" and "ExpShipI" for exposure to ship during trial and intertrial separately
\item Converting column "Posixct" to POSIXct format for date and time
\end{itemize}

@@

```{=tex}
We thus decide to filter by setting a threshold of $20$ km/h for horizontal speed.
```

@@

ggsave(path=plots_folder,filename="all_narwhals_tracks.pdf", 
       plot = map,width=10,height=5)

@@

scoresby_sound_utm=st_transform(scoresby_sound_utm,crs= "+init=EPSG:32626 +units=km")
```

@@

```{r plot coastline and land}
plot_land=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey",col="brown")
```

@@

Some points that appear on land on the map don't have distance to shore 0. This may be due to the map inaccuracy.

@@

cropped_data=dataprep[dataprep$time>12,]
pExpShip=ggplot(cropped_data,aes(time,ExpShip,col=ID))+geom_line(color="black")+
  facet_grid(ID~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  xlab("Time (h)")+ylab("Exposure to ship")+
   theme(
    axis.text = element_text(size = 12),   # Increase tick size
    axis.title = element_text(size = 16)  # Increase label size
  )

ggsave(path=plots_folder,"realExpShip_through_time.pdf", plot=pExpShip)
```
